---
title: "A Right Restricted"
subtitle: "Partisanship, Racial Backlash, and Restrictions to the Ballot Box"
thanks: I would like to thank my colleagues at the Brennan Center for Justice for their help and advice on this project. In particular, I thank Bela Tringali, Stuart Baum, Derek Paulhus for their tireless work on the Voting Laws Roundup, as well as that of Jasleen Singh, Ben Chang, and Sam Linn. Paul Gronke, John Curiel, and participants at the 2022 Annual Meeting of the Midwest Political Science Association also contributed mightily to this project. All errors are my responsibility.
author: Kevin Morris^[Researcher, Brennan Center for Justice (kevin.morris[at]nyu.edu).]
institute: "Brennan Center for Justice"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "bad_bills.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
    - \usepackage{lscape}
    - \usepackage{pdfpages}
    - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
    - \usepackage{lineno}
    # - \linenumbers
abstract: |
  In the aftermath of the United States’ 2020 presidential election, state legislatures have introduced and passed an unprecedented wave of restrictive voting bills. This project explores the state- and district-level drivers of these bills. I find that Republican-dominated states with large nonwhite populations were by far the most likely to introduce and pass these bills, while competitiveness provides little explanatory power. Meanwhile, the whitest state legislative districts in the least-white states were the most likely to be represented by lawmakers who sponsored restrictive bills, as were districts with the most racially-resentful white residents. The districts most supportive of Trump---not electorally competitive districts---were also the most likely to be represented by these lawmakers. I conclude that racial animus, the social geography of threat, and partisan signaling explain the sponsorship of restrictive voting laws, racing important normative questions about support for democracy in the United States.
---
\pagenumbering{gobble}
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(kableExtra)
library(extrafont)
```
\pagenumbering{arabic}
\doublespacing

# Introduction

On May 7, 2021, Texas legislators in the state's House of Representatives debated and passed Senate Bill 7, an omnibus bill restricting voting in various ways. The bill would reduce access to mail voting, ban drive-through and 24-hour voting, and require large counties to redistribute their polling places away from Black and Latino neighborhoods [@Ura2021a; @Ura2021]. Although this particular bill would fail after Democratic legislators broke quorum before the final vote by leaving the state, many of these provisions would ultimately become law as part of Senate Bill 1 during a special session called by the governor [@Ura2021b].

The debate in the House, however, was marked by an argument about a single phrase, used in the opening text of the bill. Senate Bill 7's self-described purpose was to "detect and punish fraud and preserve the purity of the ballot box" (§1.02). This phrase---"the purity of the ballot box"---has a long history in Texas, enshrined in the state's 1876 Constitution and used to defend the state's white primary that effectively shut nonwhites out of the political process for decades [@Knowles2021; @Morris2021b]. Democratic representative Rafael Anchía questioned the bill author's use of this "specific set of words that has a lot of meaning in state history," [quoted in @Knowles2021] saying the constitutional provision "was drafted specifically to disenfranchise Black people." The implication was clear: Texas legislators in 2021 were tapping into long-standing legal racism to pass new legislation that would disproportionately impact voters of color. The phrase was dropped from the final version of Senate Bill 1, passed in August, 2021.

The twin features of the introduction to Texas' Senate Bill 7---protection against fraud and appeals to purity---typified Republicans' rhetoric during the 2021 legislative session around the country. After losing his re-election bid in November, then-president Donald Trump claimed repeatedly that the election had been stolen [@Dale2020], a claim he has continued to maintain and that some 70% of self-identified Republicans believed by early 2022 [@Cuthbert2022]. Many state legislators also justified their support for restrictive legislation in terms similar to Oklahoma State Representative Sean Roberts (sponsor of the restrictive HB 2842 and HB 2847), who told reporters that "[I]t was very clear that the election was stolen from President Trump. We must do everything we can to close those loopholes" [quoted in @May2022]. Concerns about election security were not limited to state legislators: 147 Congressional Republicans voted against the certification of the 2020 presidential election, including a supermajority of Republicans in the House [@Yourish2021].

Despite these widespread claims, no evidence of fraud arose following the 2020 election. As the *New York Times* explained: "After bringing some 60 lawsuits, and even offering financial incentive for information about fraud, Mr. Trump and his allies have failed to prove definitively any case of illegal voting on behalf of their opponent in court---not a single case of an undocumented immigrant casting a ballot, a citizen double voting, nor any credible evidence that legions of the voting dead gave Mr. Biden a victory that wasn’t his" [@Rutenberg2020]. The lack of evidence of fraud was used to insinuate that this restrictive legislation is driven by racial animus [e.g., @BaconJr.2022].

This project relies on a comprehensive survey of voting-related bills introduced around the country in 2021 systematically collected by the Brennan Center for Justice to better understand what explains the introduction and passage of restrictive voting legislation.^[See https://www.brennancenter.org/our-work/research-reports/voting-laws-roundup-december-2021.] After detailing the relationship between state characteristics and the introduction and passage of these bills, I move to the legislative district level. I examine the characteristics of the districts from which sponsors of restrictive bills hail, asking whether the patterns can be understood as an attempt to maintain partisan control as some scholars argue [e.g., @Hicks2015] or an attempt to specifically undermine the power of racial and ethnic minorities, as others posit [e.g., @Bentele2013]. This project marks the first attempt to understand how local---and not simply state---characteristics shape support for restrictive voting laws.

The results are unequivocal: while virtually all restrictive legislative activity occured in Republican-dominated states, there are key drivers *within* these states: namely, only Republican states with *large nonwhite populations* introduce and pass restrictive provisions. Whiter states under unified Republican control introduced and passed these bills at far lower rates. Moreover, it was representatives from the whitest districts in the least-white states that were the most likely to sponsor restrictive legislation; so too were lawmakers from districts where white residents have high levels of racial resentment. Partisanship also plays a role, but not in the way past literature would suggest: sponsorship is not concentrated in contested, competitive districts where lawmakers are seeking an electoral advantage, but rather in the safest Republican districts. In short, the legislative activity related to restrictive voting bills in 2021 seems best explained by racial threat and white backlash. 

# Recent Work on Restrictive voting Laws

Over the past 15 years, scholars have explored the introduction and passage of restrictive of restrictive voting laws across the country. This work has largely focused on state-level factors, with a general consensus that these laws find the most fertile ground in states with large demographic change and a growing nonwhite electorate [@Bentele2013], where large numbers of Black Americans reside [@Behrens2003], and in electorally competitive states where Republicans hold a slight edge [@Hicks2015].

@Behrens2003 uses a historical approach to understand the passage of laws disenfranchising citizens convicted of felony offenses. As they note, all but two American states restrict voting rights for at least some incarcerated citizens; the two that do not---Maine and Vermont---are also the two whitest states in the nation. Behrens and colleagues document the rise of these restrictive laws in the aftermath of the passage of the 14th and 15th Amendments, expanding formal citizenship and granting voting rights to Black men. Drawing on @Blumer1958 and other scholars of group threat, they argue that white (male) Americans were threatened by the prospect that their sole control over the political domain was no longer so secure. Of course, their claims to racial political dominance were threatened proportionate to the number of nonwhite potential voters; as such, states with larger nonwhite populations had political incentives to develop new ways to disenfranchise Black men. They find strong support for the theory that the widespread adoption of felony disenfranchisement rules rose from this threat. "Our key finding can be summarized concisely and forcefully", they write. "The racial composition of state prisons is firmly associated with the adoption of state felon disenfranchisement laws. States with greater nonwhite prison populations have been more likely to ban convicted felons from voting than states with proportionally fewer nonwhites in the criminal justice system" [@Behrens2003, 596]. Their conclusions have been corroborated more recently. @Eubank2022 finds that states subject to strict federal oversight under the 1965 Voting Rights Act's Section 5 selectively increased the incarceration of Black Americans, providing further evidence that increased political opportunity for racial minorities leads white majorities to seek other ways of restricting their effective power.

Of course, the incarceration of citizens and subsequent legal disenfranchisement is perhaps only the most drastic example of curtailing access to the ballot.^[It bears noting, however, that being drastic does not mean it is uncommon. More than 6% of Black Americans were legally disenfranchised in 2020 due to a felony conviction. This number topped 10% in 7 of the 33 states where the Black voting age population exceeded 100,000 [@Uggen2020].] Might less extreme attempts to limit the pool of eligible voters follow a similar pattern? And do such considerations structure legislative behavior in the modern era? @Bentele2013 consider the introduction and passage of 5 types of restrictive voting legislation ("photo identification requirements, proof of citizenship requirements, laws that introduce restrictions on voter registration, restrictions on absentee and early voting, and restrictions on participation by felons" (1095)) over the 2006--2011 period. They conclude that the strongest predictor of the introduction and passage of restrictive voting laws is the political power demonstrated by racial and ethnic minorities, arguing that "legislative developments in this policy area remain heavily shaped by racial considerations" [@Bentele2013, 1089]. At the same time, they find no evidence that prevalence of voter fraud impacted the introduction of restrictive provisions and that it was "only a minor contributing factor" to the passage of these laws in 2011 (1103).

A further insight from @Bentele2013---that restrictive provisions are passed most frequently in electorally-competitive states---is corroborated by @Hicks2015. Looking specifically at the introduction and passage of restrictive voter identification laws in the early 2000s. They find that states with more Republican legislators were considerably more likely to enact these provisions---but that "Republicans have not pursued this scorched-earth policy in all states, nor have they done so consistently over time" (29--30). Instead, Republicans were more likely to pass these bills where their electoral majorities were slim. Hicks et al. [-@Hicks2015, 18] thus conclude that "where elections are competitive, the furtherance of restrictive voter ID laws is a means of maintaining Republican support while curtailing Democratic electoral gains." Other work [e.g., @Biggers2015; @Wang2012] also indicates that restrictive voting laws are passed by Republican-dominated legislatures to shore up flagging electoral majorities.

This scholarship sheds important light on where restrictive voting laws are the most likely to go into effect, and the results are not encouraging. There is strong evidence that racial threat predicts the passage of these restrictive bills across the country, even as legislators proclaim that the changes are needed to combat widespread fraud [see, for instance, @Piven2009; @Minnite2010]. Important as this research has been, however, it fails to explain the full set of dynamics between demographic composition and bill introduction. The explosion in the introduction of restrictive voting laws in 2021 makes this clear: according to the data from the Brennan Center for Justice used throughout this project, lawmakers in every state except Vermont introduced at least one voting bill in 2021 containing restrictive voting provisions. Moreover, the number of restrictive provisions introduced and passed in 2021 has little historical precedent: 880 restrictive provisions were introduced and 93 were passed. By way of comparison, @Bentele2013---which also used data from the Brennan Center---calls roughly the roughly 20 passed provisions in 2011 a "dramatic increase" (1088; see their Figure 2).

Clearly, something more complex than state-level factors are at play in the contemporary push to restrict voting rights. By considering not only state-level factors but also examining the demographics of the districts represented by legislators who introduce, co-sponsor, and vote for these restrictive bills, this project marks a significant step forward in understanding how racial threat's influence on the policy-making process is mediated by factors at multiple political levels. The following section steps back to engage with the (racial) threat literature and, more specifically, consider how spatially-situated theories of threat help us to formulate expectations about the roles played by state and local factors in the introduction and passage of restrictive voting laws.

# A Changing Electorate and Threat

Scholars across the social sciences have long noted the importance of threat to the policy-making process; indeed, each of the studies discussed in the previous section implicitly or explicitly draw on this literature. @Tilly1978 separates collective action into three categories: defensive, offensive, and preparatory (73). Of these, two---defensive and preparatory---are explicitly linked to threats, where political actors pool their resources to fend of challenges to their interests, or to regain what has already been lost. @Beck2000 extends this theory to note that defensive actors need only *perceive* that their interests have been compromised to mobilize in a reactionary way; the *reality* of any worsened station is perhaps less important. These threats can take multiple forms, be they economic, political, or demographic [@VanDyke2002].

In recent years, increasing attention has been paid to how different levels of spatial organization and threat can interact with one another [@Zhang2018]. @Tilly2015 explains how social movements can undergo what they call an "upward scale shift," which they say "moves contention beyond its local origins, touches on the interests and values of new actors, involves a shift of venue to sites where contention may be more or less successful, and can threaten other actors or entire regimes" (125). In other words, political actors may move beyond the local context to make use of institutional tools available only at higher levels of government.

Although scholars have recognized the importance of space and scale in political and legislative activity, relatively less attention has been paid to how the *interaction* of conditions at these different levels plays out. A notable exception to this is @Andrews2015, which explores how racial threat, group contact, and differential levels of government structure legislative activity. They argue that "threat arises primarily from interactions between spatially proximate units at the local level... and therefore higher-level policy change at the state level is not reducible to the variables driving local policy" (476). In other words, examining local and state characteristics alone is not sufficient to understand legislator support for racially conservative policy changes; instead, responses to racial threat arise from the *interaction* of these circumstances.

Threat is clearly a major driver of policy in the United States. As the research on voting laws makes clear, states pass more restrictive legislation when there are more racial minorities whose political power threatens the established power structure. However, sociological models of mobilization, scale shifts, and geographical interplay pushes us to think more seriously about precisely where the support for racially-restrictive legislation comes from. On the one hand, we might expect that racially-diverse localities where whites maintain a political edge would feel the most threatened by nonwhite voters, leading their representatives to support more restrictive legislation. On the other hand, white and homogeneous areas of diverse states might be threatened by rising political power elsewhere in their own states, and their homogeneity might provide the basis for coordinated pressure on their representatives. Identifying the source of legislative support for racially-restrictive policymaking *within* states is of key importance for better understanding the geographical / political topography of racial threat.

# Methods and Expectations

Throughout my analyses, I rely on the Voting Laws Roundup, a project of the Brennan Center for Justice at NYU School of Law. The Brennan Center systematically reviews all laws introduced around the country that relate to voting and the administration of elections in each state. The Brennan Center identifies these bills using string-searches in Westlaw, and then separates each bill introduced into its constituent provisions, using two coders to designate each provision as "restrictive," "neutral," or "expansive." Each provision is also assigned to a category describing its content (such as "voter ID", "polling place count", or "funding for poll workers"). Each bill's provisions are identified when a bill first includes provisions related to voting, and updated if a bill is passed. In other words, if a bill is introduced with some voting provisions, is subsequently amended to include other voting provisions, but ultimately fails to pass, only the original provisions are included. Figure \ref{fig:cols} shows the categorical breakdown of restrictive provisions introduced and passed, while Figure \ref{fig:maps} shows the geographical distribution of these provisions.

```{r cols, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 5, fig.cap = "\\label{fig:cols}Categories of Restrictive Provisions, 2021", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/column_cat_provs.rds"))
```

```{r maps, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 3, fig.cap = "\\label{fig:maps}Restrictive Provisions, 2021", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/map_provs.rds"))
```

I control for the partisan control of each state in two ways. Following @Hicks2015, it seems possible that electorally competitive states where Republicans hold unified power would be most likely to introduce and pass bad provisions. I thus include 2 dummies: one measuring whether the state was competitive (that is, Biden received between 45% and 55% of votes), and one measuring whether Republicans held unified control in 2021. Data on electoral competitiveness comes from the MIT Election Data and Science Lab [-@MITElectionDataandScienceLab2021], and data on partisan control comes from the National Conference of State Legislatures^[See https://www.ncsl.org/documents/elections/Legis_Control_2-2021.pdf.] Although Nebraska's unicameral state legislature is formally nonpartisan, they are considered to be under unified Republican control for the purposes of this study.

The analyses conducted at the level of the legislative district primarily leverage data from LegiScan, an organization that tracks state-level bills around the nation. Importantly, LegiScan records the legislators who sponsor each of these bills. By merging the bills including restrictive provisions with the LegiScan data, I identify all districts represented by a lawmaker who sponsored at least one restrictive voting law.

The primary independent variables for the first set of district-level analyses are the white share of the district *and* the white share of the state. These are included to test whether the influence of the whiteness of a district on the probability that a lawmaker sponsors a restrictive bill is influenced by state-level factors. I also test the relationship between sponsorship and partisanship at the district level by controlling for the share of the district won by Donald Trump in the 2020 presidential election. I estimate Trump's two-party vote share in each district by aggregating up from precinct-level results published by the Voting Election and Science Team [-@VotingandElectionScienceTeam2022]. I assign each precinct to the upper- and lower-chamber district in which its geographical center is located.^[While this will not perfectly estimate voteshare in chambers where precincts cross district lines, there is little reason to expect this will systematically distort voteshare estimates.] This coverage is not perfect: in Kentucky and West Virginia, where precinct-level results are not available, I use population-weighted county-level results. Specifically, I assign each Census block the Trump vote share of its home county. District vote share is calculated as the population-weighted mean of Trump vote share in each block in the district.

Of course, administrative and demographic data cannot give us insight into the political disposition of district residents. As such, I also incorporate survey data from the 2020 Cooperative Election Study to test whether districts' racial resentment scores are associated with the sponsorship of restrictive bills. The 2020 CES asks white voters how strongly they agree (on a scale of 1 to 5) with two statements related to racial resentment: *Irish, Italians, Jewish and many other minorities overcame prejudice and worked their way up. Blacks should do the same without any special favors* and *Generations of slavery and discrimination have created conditions that make it difficult for blacks to work their way out of the lower class*. I reverse code agreement with the first statement, such that higher scores for both statements are associated with higher levels of racial resentment. Respondents' racial resentment scores are calculated as the mean of their response to these two questions. I retain only the responses of white respondents.

While the CES data do not include respondents' home legislative districts, the survey makes home ZIP codes available. To calculate district resentment scores, I begin by assigning every Census block in the country the mean resentment score of the ZIP code in which its centroid falls. District resentment scores are then calculated as the population-weighted average racial resentment score of all blocks in the district.

In addition to these measures of voting legislation and policy environment, I incorporate demographic data at the state and legislative-district level from the Census Bureau's American Communities Survey. These estimates are the 5-year numbers ending with 2020. I also incorporate information about how difficult voting was prior to 2021 using the Cost of Voting Index (COVI) [@Schraufnagel2020].

Based on the literature discussed above, I pose two hypotheses at the state level:

* **H1a**: Restrictive voting legislation is best explained by theories of racial threat and white backlash. Therefore, other things being equal, states where whites make up a smaller share of the population were more likely to introduce and pass restrictive voting provisions. I expect this to be especially true in states with unified Republican control---that is, in more racially diverse states where the conservative party can pass restrictive bills into law.

* **H1b**: Restrictive voting legislation is best explained by theories of partisan advantage. Therefore, other things being equal, electorally-competitive states with unified Republican control were the most likely to introduce and pass restrictive provisions.

After presenting tests of these hypotheses at the state-level, I move to analyses at the level of the legislative chamber. The empirical framework is the same, though the dependent variable changes from the *count* of restrictive provisions to a dummy variable indicating *whether* a legislator sponsored restrictive provisions, due to smaller scale.

In light of recent sociological work on the geography of racial threat [@Andrews2015], **H2**: I expect that whiter districts in less-white states were the most likely to be represented by legislators that sponsored restrictive voting provisions.

I also test the effect of partisanship and competitiveness along the lines of past scholarship at the district level. This past work would predict that **H3a**: lawmakers from competitive districts controlled by Republicans are the most likely to support restrictive voting legislation. If, however, the sponsorship of such legislation is tied more to racial threat and signalling to an aggrieved base, I predict that **H3b**: representatives from the most conservative districts were the most likely to sponsor these bills.

Finally, I test the relationship between whites' racial resentment and lawmaker sponsorship of restrictive bills, hypothesizing that **H4**: Districts where white voters had higher levels of racial resentment were more likely to be represented by legislators that sponsored restrictive voting provisions.

# Results

## State-Level Legislative Behavior

In Table \ref{tab:state}, I present the results of tests on my first set of hypotheses. In Models 1--3, the dependent variable is the count of restrictive provisions introduced; in Models 4--6, it is the number of restrictive provisions passed. In each case, I use a robust regression (using `rlm()` in `R` from the `MASS` package) to prevent outlier states like Texas from exerting too much influence on the models. As I show in the SI, these results are substantively similar using OLS.

\newpage
\begin{landscape}
\input{"../temp/state_reg.tex"}
\end{landscape}
\newpage

Because interaction coefficients can be difficult to interpret in table format, I here present the predicted probabilities of counts of introduced and passed provisions in Figure \ref{fig:mefs-state}. These results are presented with and without covariates to visually demonstrate how their inclusion alters the substantive relationships of interest. The top panels plot the results from Models 1 and 3 from Table \ref{tab:state}; the bottom panels present Models 4 and 6. In Panels (C) and (D), other covariates are held at their means.

```{r mefs-state, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 5, fig.cap = "\\label{fig:mefs-state}Restrictive Provisions Introduced and Passed, by Race and 2020 COVI", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/mef_race_both.rds"))
```

We can easily see from Table \ref{tab:state} and Figure \ref{fig:mefs-state} that the wave of restrictive provisions in 2021---both those introduced and those passed---have much to do with partisanship *and* race. States that were not under unified Republican control neither introduced nor passed restrictive voting provisions in appreciable numbers. Among states with unified Republican control, however, race appears to be a driving factor: even after controlling for competitiveness in the states, less-white Republican-dominated states introduced and passed far more restrictive provisions than whiter states with unified Republican control. This is perhaps unsurprising: Texas, Georgia, Florida, and Arizona are all fully controlled by the Republican party, and each of these states passed either an omnibus bill or a series of restrictive laws in 2021. Moreover, as a comparison of Models 1 and 3, and 4 and 6 indicate, the relationships between state control, race, and restrictive provisions are not highly influenced by the inclusion of additional covariates. The coefficients on the independent variables of interest do not move very much. This provides strong support for **H1a**: these state-level patterns are consistent with theories of racial backlash, where the conservative party responds to nonwhite populations by restricting voting when it controls the levers of government.

Table \ref{tab:state} also provides some insight into the relationship between competitiveness, Republican domination, and restrictive provisions. In the naive models (2 and 5), few of the coefficients are significant at the 95% confidence level. These models indicate that perhaps competitive states were more likely to *introduce* more restrictive provisions, though partisan control does not explain differences among competitive states. Further, competitive states with unified Republican control were perhaps more likely to pass restrictive provisions than other competitive states. After accounting for the other covariates in the models---including racial characteristics interacted with Republican control---, it seems that more restrictive provisions were introduced (but not passed) in states with unified Republican control. Competitive states with unified Republican control passed more restrictive provisions, ceteris paribus, though the negative (statistically non-significant) coefficient on competitiveness in other states reduces the substantive importance of this relationship. All told, there is little evidence that strongly supports **H1b**.

In short: theories of racial threat and white backlash do a far better job of explaining the observed patterns in the 2021 legislative context than do ones accounting only for partisan control and competitiveness.

## District-Level Legislative Behavior

Having determined that the geographic concentration of the restrictive voting bills in 2021 was highly consistent with the racial threat theory, I turn to the sub-state level to better understand these processes. As discussed above, I expect to find evidence of backlash against racial threat in the whitest parts of the least white states, and where white Americans are more racially resentful.

Table \ref{tab:chamb} presents the results of OLS regressions^[As I show in the SI, the results presented throughout these analyses are generally robust to using logistic regression instead of OLS.] investigating the likelihood that a legislator from a given district sponsored at least one restrictive voting provision in 2021. About 22% of lower chamber legislators and 20% of upper chamber legislators sponsored at least one bill with a restrictive provision.

\input{"../temp/district_reg.tex"}

As before, I plot the predicted probability of restrictive bill sponsorship to aid in the interpretation of the substantive relationships. Figure \ref{fig:mef-dis} presents like predicted likelihood of the sponsorship of a provision at different levels of district- and state-level racial compositions. In the first row, I present the results of Models 1 and 3, where only racial characteristics are included. The second row plots the models that include covariates. Once again, all other covariates are held at their means.

```{r mefs-cha, echo = FALSE, message = F, warning = F, fig.width = 6, fig.height = 7, fig.cap = "\\label{fig:mef-dis}Sponsorship and Race", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/mef_chamb_good.rds"))
```

Table \ref{tab:chamb} and Figure \ref{fig:mef-dis} show that the relationship between district-level demographics and restrictive voting sponsorship are highly moderated by state-level characteristics, as expected. Namely, representatives from the whitest legislative districts in the most racially diverse states were by far the most likely to sponsor restrictive legislation. There is little relationship between district-level demographics and the probability of sponsoring a restrictive voting bill in the whitest states; it is only in the least-white states that this relationship becomes apparent. Although these relationships are moderated slightly with the inclusion of relevant sociodemographic and political covariates, the patterns remain clear: representatives from white districts in states with large nonwhite populations---whose (growing) demographic and political power could inspire fear in the more homogeneously-white parts of the state---were disproportionately likely to sponsor bills with provisions making voting more difficult.

Table \ref{tab:chamb} (and the predicted probabilities plotted in Figure \ref{fig:mef-comp}) also indicates that the sponsorship of restrictive voting laws at the district-level follows substantially different patterns than those established previously looking at state-level factors. As discussed above, electorally-competitive *states* are the most likely to pass restrictive voting laws in an attempt to maintain control. **H3a** extended this logic to the district level, hypothesizing that lawmakers from competitive districts would act similarly. The results, however, strongly support **H3b**, indicating lawmakers from the most Republican districts were the most likely to sponsor these bills. If these bills were being supported out of genuine fear about fraud distorting the results in close elections, we would have expected to see them supported most by precisely the lawmakers whose districts were competitive. Instead, representatives from *safe* districts who could not reasonably fear electoral loss in a general election from fraud---even if it existed---are the most common sponsors. While partisanship plays a role in these sponsorships, *something other than electoral competitiveness* drives the Republicans who sponsor these provisions. As discussed above, this is likely racial threat among white Republicans.

```{r mefs-comp, echo = FALSE, message = F, warning = F, fig.width = 6, fig.height = 7, fig.cap = "\\label{fig:mef-comp}Sponsorship and Partisanship", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/mef_chamb_party.rds"))
```

To test whether racial resentment really is associated with the sponsorship of restrictive bills, I turn to the CES data. After estimating each district's resentment score, I ask whether this score is associated with the probability that a lawmaker sponsored a bill with at least one restrictive provision. Figure \ref{fig:mef-rr} plots the predicted probabilities of sponsorship, both before and after controlling for other covariates (the regression table can be found in the SI). The top panels demonstrate a strong bivariate relationship between racial resentment and sponsorship: districts where white respondents to the CES were more racially resentful were far more likely to be represented by a lawmaker who sponsored a restrictive bill. This is especially true in the upper chamber.

```{r mefs-rr, echo = FALSE, message = F, warning = F, fig.width = 6, fig.height = 7, fig.cap = "\\label{fig:mef-rr}Sponsorship and Racial Resentment", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/mef_rr.rds"))
```

These relationships are moderated by the inclusion of sociodemographic controls, especially the partisan measures. Although these relationships are smaller, they nevertheless remain substantively quite large and statistically significant. In the lower chamber, the most-resentful districts were more than fifty percent (10 points) more likely to be represented by a lawmaker sponsoring a restrictive bill than the least-resent districts, other things being equal. In the upper chamber, the probability of being represented by a lawmaker sponsoring a restrictive bill for the most-resentful districts was *double* that of the least resentful districts. The explanatory power of racial resentment above-and-beyond what can be explained by racial and partisan measures is striking and provides exceedingly strong evidence for **H4**.

# Discussion

While legislators claim to pass restrictive voter policies under the guise of improved security, the data unambiguously points to white backlash to perceived racial threat. While discussing what she calls the "political work of fraud allegations," Lorraine Minnite argued a decade ago that claims of fraud could be explained by "the existence of marginalized subjects within the political culture whose presence alone stands in as the evidence of the alleged fraud" [@Minnite2010, 87]. In other words, the rhetoric of voter fraud draws boundaries around who does and does not count as a citizen, and whose political participation is inherently suspect.

A similar pattern emerges in this study of regressive voting laws introduced and passed in 2021, following the hard-fought and deeply polarizing 2020 presidential election and subsequent violent attempt to overturn the results on January 6th, 2021. Although the push to restrict voting access was extremely broad in 2021---more than 1 in 3 Americans lived in a district represented by a legislator who sponsored at least one restrictive provisions---the push was concentrated in states and legislative districts with particular, systematic demographic characteristics.

As anti-democratic forces continue to undermine faith in the United States' electoral infrastructure, understanding the source of this backlash is of key importance. I uncover no evidence that lawmakers from competitive districts were more likely to sponsor restrictive bills. In a year with such widespread restrictive voting legislation, why is support for these measures not concentrated in the districts where shifting the composition of the electorate (or preventing fraud) by a tiny amount might matter? The explanation for this behavior might lie in theories of legislator signalling. As @Rocca2010 note, "members of an attentive public pay close attention to legislators' stances. Representatives know this and use non–roll call forums to signal attentive groups that they are 'on their side.' Groups, in turn, reward representatives sympathetic to their causes with campaign contributions" (387). Legislators from less polarized districts, on the other hand, are more likely to take moderate positions, reflecting the influence of a competitive electorate [@Kirkland2014]. Future work should directly interrogate whether legislators who signal support for restrictive voting rights gain campaign such benefits as campaign contributions or increased turnout.

This research demonstrates that the backlash is inextricably linked with race: insofar as sponsorship proxies a district's appetite for restrictive voting legislation, these bills found the most fertile soil where white voters are concentrated in racially diverse states. The incorporation of survey data makes the relationships even clearer, showing that racial resentment among white Americans has explanatory power above-and-beyond any partisan effect. Similarly, the heterogeneity *within* states with unified Republican control sheds new light on how we understand the introduction and passage of these bills. In 2021, it was the nonwhite Republican states---and not in any clear-cut way the electorally-competitive Republican states---that were the most active in this policy space. All told, race plays a clear and unambiguous role in the restrictive wave of voting legislative that has characterized the post-2020 legislative landscape.

\newpage
# References {-}
