---
title: "A Right Restricted"
subtitle: "Understanding the Introduction and Passage of Restrictive Voting Laws"
thanks: TKTKT
author: Kevin Morris^[Researcher, Brennan Center for Justice (kevin.morris[at]nyu.edu).]
institute: "Brennan Center for Justice"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  bookdown::pdf_document2:
    toc: no
    fig_caption: yes
    latex_engine: xelatex
    keep_tex: true
bibliography: "bad_bills.bib"
link-citations: yes
fontsize: 12pt
header-includes:
    - \usepackage{rotating}
    - \usepackage{setspace}
    - \usepackage{lscape}
    - \usepackage{pdfpages}
    - \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
    - \usepackage{lineno}
    # - \linenumbers
abstract: |
  TKTKT
---
\pagenumbering{gobble}
\pagebreak
```{r setup, echo = F, message = F, warning = F}
knitr::opts_knit$set(root.dir = "..", dev = "cairo_pdf")
library(kableExtra)
library(extrafont)
```
\pagenumbering{arabic}
\doublespacing

# Introduction

On May 7, 2021, Texas legislators in the state's House of Representatives debated and passed Senate Bill 7, an omnibus bill restricting voting in various ways. The bill would reduce access to mail voting, ban drive-through and 24-hour voting, and require large counties to redistribute their polling places away from Black and Latino neighborhoods [@Ura2021a; @Ura2021]. Although this particular bill would ultimately fail after Democratic legislators broke quorum before the final vote by leaving the state, many of these provisions would ultimately become law as part of Senate Bill 1 during a special session called by the governor [@Ura2021b].

The debate in the House, however, was marked by an argument about a single phrase, used in the opening text of the bill. Senate Bill 7's self-described purpose was to "detect and punish fraud and preserve the purity of the ballot box" (§1.02). This phrase---"the purity of the ballot box"---has a long history in Texas, enshrined in the state's 1876 Constitution and used to defend the state's white primary that effectively shut nonwhites out of the political process for decades [@Knowles2021; @Morris2021b]. Democratic representative Rafael Anchía questioned the bill author's use of this "specific set of words that has a lot of meaning in state history," [quoted in @Knowles2021] saying the constitutional provision "was drafted specifically to disenfranchise Black people." The implication was clear: Texas legislators in 2021 were tapping into long-standing legal racism to pass new legislation that would disproportionately impact voters of color. The phrase was dropped from the final version of Senate Bill 1, passed in August.

The twin features of the introduction to Texas' Senate Bill 7---protection against fraud and appeals to purity---typified the 2021 legislative session around the country. After losing his re-election bid in November, then-president Donald Trump claimed repeatedly that the election had been stolen [@Dale2020], a claim he has continued to maintain and that some 70% of registered Republicans believed by early 2022 [@Cuthbert2022]. Despite these widespread beliefs, no evidence of fraud arose following the 2020 election, as the New York Times explained: "After bringing some 60 lawsuits, and even offering financial incentive for information about fraud, Mr. Trump and his allies have failed to prove definitively any case of illegal voting on behalf of their opponent in court---not a single case of an undocumented immigrant casting a ballot, a citizen double voting, nor any credible evidence that legions of the voting dead gave Mr. Biden a victory that wasn’t his" [@Rutenberg2020]. By pointing to the fear of fraud, however, legislators around the country enacted voting restrictions so wide that they potentially amounted to "the most sweeping contraction of ballot access in the United States since the end of Reconstruction" [@Gardner2021b].

This project relies on a comprehensive survey of voting-related bills introduced around the country in 2021 systematically collected by the Brennan Center for Justice.^[See https://www.brennancenter.org/our-work/research-reports/voting-laws-roundup-december-2021. Provision-level data available from the Brennan Center upon request.] I examine proposed and enacted legislation in light of already-existing voting regimes to test whether more restrictive provisions were considered in states with more permissive voting regimes, providing evidence of "system hardening" in places with more lenient voting environments. I also explore the relationship between these provisions and racial demographics at the state and legislative-district level to test how well theories of racial threat explain the observed legislative activity. The results are unequivocal: states with large nonwhite populations where voting was *already* restricted made it even more difficult to vote in 2021. Moreover, it was representatives from the whitest districts in the least-white states---that is, the districts most theoretically likely to exhibit behavior explained by racial threat---that were the most likely to sponsor restrictive legislation. Expansive provisions followed a decidedly different pattern, uniformly more likely to be sponsored by representatives from less-white districts.

# Recent Work on Restrictive voting Laws

Over the past 15 years, scholars have explored the introduction and passage of restrictive of restrictive voting laws across the country. This work has largely focused on state-level factors, with a general consensus that these laws find the most fertile ground in states with large demographic change and a growing nonwhite electorate [@Bentele2013], where large numbers of Black Americans reside [@Behrens2003], and in electorally competitive states where Republicans hold a slight edge [@Hicks2015].

@Behrens2003 uses a historical approach to understand the passage of laws disenfranchising citizens convicted of felony offenses. As they note, all but two American states restrict voting rights for at least some incarcerated citizens; the two that do not---Maine and Vermont---are also the two whitest states in the nation. Behrens and colleagues document the rise of these restrictive laws in the aftermath of the passage of the 14th and 15th Amendments, expanding formal citizenship and granting voting rights to Black men. Drawing on @Blumer1958 and other scholars of group threat, they argue that white (male) Americans were threatened by the prospect that their sole control over the political domain was no longer so secure. Of course, their claims to racial political dominance were threatened proportionate to the number of nonwhite potential voters; as such, states with larger nonwhite populations had political incentives to develop new ways to disenfranchise Black men. They find strong support for the theory that the widespread adoption of felony disenfranchisement rules rose from this threat. "Our key finding can be summarized concisely and forcefully", they write. "The racial composition of state prisons is firmly associated with the adoption of state felon disenfranchisement laws. States with greater nonwhite prison populations have been more likely to ban convicted felons from voting than states with proportionally fewer nonwhites in the criminal justice system" [@Behrens2003, 596]. Their conclusions have been corroborated more recently. @Eubank2022 finds that states subject to strict federal oversight under the 1965 Voting Rights Act's Section 5 selectively increased the incarceration of Black Americans, providing further evidence that increased political opportunity for racial minorities leads white majorities to seek other ways of restricting their effective power.

Of course, the incarceration of citizens and subsequent legal disenfranchisement is perhaps only the most drastic example of curtailing access to the ballot.^[It bears noting, however, that being drastic does not mean it is uncommon. More than 6% of Black Americans were legally disenfranchised in 2020 due to a felony conviction. This number topped 10% in 7 of the 33 states where the Black voting age population exceeded 100,000 [@Uggen2020].] Might less extreme attempts to limit the pool of eligible voters follow a similar pattern? And do such considerations structure legislative behavior in the modern era? @Bentele2013 consider the introduction and passage of 5 types of restrictive voting legislation ("photo identification requirements, proof of citizenship requirements, laws that introduce restrictions on voter registration, restrictions on absentee and early voting, and restrictions on participation by felons" (1095)) over the 2006--2011 period. They conclude that the strongest predictor of the introduction and passage of restrictive voting laws is the political power demonstrated by racial and ethnic minorities, arguing that "legislative developments in this policy area remain heavily shaped by racial considerations" [@Bentele2013, 1089]. At the same time, they find no evidence that prevalence of voter fraud impacted the introduction of restrictive provisions and that it is "only a minor contributing factor" to the passage of these laws in 2011 (1103).

A further insight from @Bentele2013---that restrictive provisions are passed most frequently in electorally-competitive states---is corroborated by @Hicks2015. EXPAND

This scholarship sheds important light on where restrictive voting laws are the most likely to go into effect, and the results are not encouraging. There is strong evidence that racial threat predicts the passage of these restrictive bills across the country, even as legislators proclaim that the changes are needed to combat widespread fraud [see, for instance, @Piven2009; @Minnite2010]. Important as this research has been, however, it fails to explain the full set of dynamics between demographic composition and bill introduction. The explosion in the introduction of restrictive voting laws in 2021 makes this clear: according to the data from the Brennan Center for Justice used throughout this project, just one state (Vermont) introduced *no* voting bills in 2021 containing no restrictive provisions. Moreover, the number of restrictive provisions introduced and passed in 2021 has little historical precedent: 880 restrictive provisions were introduced and 93 were passed. By way of comparison, @Bentele2013---which also used data from the Brennan Center---calls roughly the roughly 20 passed provisions in 2011 a "dramatic increase" (1088; see their Figure 2).

Clearly, something more complex than state-level factors are at play in the contemporary push to restrict voting rights. By considering not only state-level factors but also examining the demographics of the districts represented by legislators who introduce, co-sponsor, and vote for these restrictive bills, this project marks a significant step forward in understanding how racial threat's influence on the policy-making process is mediated by factors at multiple political levels. The following section steps back to engage with the (racial) threat literature and, more specifically, consider how spatially-situated theories of threat help us to formulate expectations about the roles played by state and local factors in the introduction and passage of restrictive voting laws.

# A Changing Electorate and Threat

Scholars across the social sciences have long noted the importance of threat to the policy-making process; indeed, each of the studies discussed in the previous section implicitly or explicitly draw on this literature. @Tilly1978 separates collective action into three categories: defensive, offensive, and preparatory (73). Of these, two---defensive and preparatory---are explicitly linked to threats, where political actors pool their resources to fend of challenges to their interests, or to regain what has already been lost. @Beck2000 extends this theory to note that defensive actors need only *perceive* that their interests have been compromised to mobilize in a reactionary way; the *reality* of any worsened station is perhaps less important. These threats can take multiple forms, be they economic, political, or demographic [@VanDyke2002].

The social sciences are replete with empirical evidence of the mobilizing effect of these three forms of threat. Early work on collective action demonstrates that unemployed workers 

Only recently, however, have the spatial dynamics of racial threat and policy threat been considered with serious attention. Andrews and Seguin [-@Andrews2015, 476] argues that "threat arises primarily from interactions between spatially proximate units at the local level... and therefore higher-level policy change at the state level is not reducible to the variables driving local policy." In other words, examining local and state characteristics alone is not sufficient to understand legislator support for racially conservative policy changes; instead, responses to racial threat arise from the *interaction* of these circumstances.

# Methods and Expectations

Throughout our analyses, I rely on the Voting Laws Roundup, a project of the Brennan Center for Justice at NYU School of Law. The Brennan Center systematically reviews all laws introduced around the country that relate to voting and the administration of elections in each state. The Brennan Center then manually separates each bill introduced into its constituent provisions, using two coders to designate each provision as "restrictive," "neutral," or "expansive." Each provision is also assigned to a category describing its effect (categories include effects such as "voter ID", "polling place count", or "funding for poll workers"). In 2021, the Brennan Center identified 2,859 bills that were introduced and related to voting, consisting of 5,927 provisions. Of these provisions, 812 were deemed restrictive; 3,445 neutral; and 1,670 expansive. A total of 72 restrictive provisions were passed into law that year; 157 neutral provisions and 19 expansive ones were also passed. Figure \ref{fig:cols} shows the categorical breakdown of restrictive provisions introduced and passed, while Figure \ref{fig:maps} shows the geographical distribution of these provisions.

```{r cols, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 5, fig.cap = "\\label{fig:cols}Categories of Restrictive Provisions, 2021", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/column_cat_provs.rds"))
```

```{r maps, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 3, fig.cap = "\\label{fig:maps}Restrictive Provisions, 2021", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/map_provs.rds"))
```

These provisions are then merged with data from LegiScan, which tracks bills of all kinds in statehouses throughout the United States. The LegiScan data includes information on bill sponsorship, roll-call votes, and dates of introduction. These data are used to identify the (co)sponsors of all the voting laws identified by the Brennan Center, as well as the legislators voting in favor or against them.

To account for pre-2021 variation in the electoral landscape, I incorporate each state's score on the (pre-COVID) 2020 Cost of Voting Index [@Schraufnagel2020] (COVI). This index captures many of the same items tracked by the Brennan Center, such as whether a state has same-day registration, voter ID laws, and the number of days of early voting. The index is widely used by social scientists to measure the difficulty of voting [e.g. @Juelich2020; @Pabayo2021; @Rackey2022]. A higher value of the COVI indicates that voting is more difficult in that state, and the 2020 COVI ranges from -2.9 in Oregon to 1.4 in New Hampshire.

In addition to these measures of voting legislation and policy environment, I incorporate demographic data at the state and legislative-district level from the Census Bureau's American Communities Survey. Except where noted, these estimates are the 5-year numbers ending with 2019.^[At the time of writing, the 2020 estimates were not yet available. The study will be updated when they become available.]

Based on the literature discussed above, I pose competing hypotheses at the state-level:

* States introduced and passed restrictive bills in 2021 in order to increase the security of their elections. Insofar as expansive voting laws are associated with (perceived) election insecurity, **H1a**: I expect that more restrictive voting provisions were introduced and passed in 2021 in states where the COVI was lower in 2020.

* According to past literature, however, the COVI might capture more than just the permissiveness of the voting regime. As discussed above, @Bentele2013 and others argue that the difficulty of voting in a given state reflects white backlash and a response to racial threat. This theory would predict an entirely different legislative pattern in 2021 than H1a. If the 2020 COVI is indicative of how likely a state is to react to displays of nonwhite political power, I would expect **H1b**: states with higher COVIs introduced and passed more restrictive legislation in 2021 than states with lower COVIs.

If I find support for **H1b**, this implies that the COVI is, at least in part, a measure of state-level reactiveness to racial threat. This further implies that less-white states with a higher 2020 COVI were more likely to respond to the 2020 election with restrictive election legislation. As such, **H2**: conditional on H1b, I expect that less-white states with a higher 2020 COVI introduced and passed more restrictive provisions in 2021 than less-white states with a lower 2020 COVI.

After presenting tests of these hypotheses at the state-level, I move to analyses at the level of the legislative chamber. The empirical framework is the same, though the dependent variable changes from the *count* of restrictive provisions to a dummy variable indicating *whether* a legislator sponsored restrictive provisions, due to smaller scale.

In light of recent sociological work on the geography of racial threat [@Andrews2015], **H3**: I expect that whiter districts in less-white states were the most likely to be represented by legislators that sponsored restrictive voting provisions.

# Results

## State-Level Legislative Behavior

In Table \ref{tab:state}, I present the results of tests on Hypotheses 1a, 1b, and 2. The coefficients on *2020 COVI* provide tests for H1a and H1b. Because the regression has just 50 observations, I mark observations that are significant at just the 90% confidence-level; while these are not statistically significant at the general threshold, they made indicate an underlying relationship.

Across the various specifications, the 2020 Cost of Voting Index is perhaps associated with an *increase* in the number of restrictive provisions introduced (though the coefficient does not generally reach typical levels of significance in Models 1, 2, and 4). Put differently: states where it was already the hardest to vote may have made it *even harder* to do so during the 2021 legislative session. This is provides support for H1b and undermines support for H1a. The fact that places where it was already hard to vote introduced and passed more restrictive legislation indicates that changes in voting legislation may bear little relationship to real concerns about election integrity.

In Models 1--4, the *p*-value of the joint significance tests for the coefficients on the *Nonhispanic White* terms is generally above 0.1, indicating a lack of a strong relationship between the racial composition of the state and its legislative behavior with respect to voting in 2021.

Models 5--8, however, indicate that the relationship between the white share of the state and the restrictive activity is *highly* moderated by the 2020 COVI. Models 5--8 mirror models 1--4, with the addition of interactive terms between these variables. In each case, the addition of the interaction terms more than doubles the adjusted R^2^; in the case of the models without other covariates, it is quadrupled. Once the interactions are included, the coefficients on the key independent variables are statistically significant at very high levels.

Because interaction and polynomial terms are difficult to interpret on their own, Figure \ref{fig:state-mef} plots the predicted number of restrictive provisions introduced and passed in 2021 at different levels of the COVI and state racial compositions. Figure \ref{fig:state-mef}(A) plots Model 6 from Table \ref{tab:state}, while \ref{fig:state-mef}(B) plots Model 8. In both cases, the other covariates are held that their means.

\newpage
\begin{landscape}
\input{"../temp/state_reg.tex"}
\end{landscape}

```{r mefs-state, echo = FALSE, message = F, warning = F, fig.width = 7.5, fig.height = 5, fig.cap = "\\label{fig:state-mef}Restrictive Provisions Introduced and Passed", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/mef_state_good.rds"))
```

Figure \ref{fig:state-mef} lays these relationships out starkly: less-white states where the COVI was already high were substantially more likely to introduce and pass restrictive legislation than A) whiter states with high COVIs, and B) less-white states with lower COVIs. Generally speaking, the whitest states did not consider restrictive legislation in 2021, regardless of their 2020 COVI. The same was true of less-white states with lower 2020 COVIs. Figure \ref{fig:state-mef} indicates that the regressive legislation of 2021 was concentrated in a handful of states and provides very strong support for H2 (and, by extension, H1b). In the Supplementary Information, I show that these relationships are very different when we consider the introduction and passage of *expansive* voting bills.

This is wholly consistent with H2 and the existing literature more generally. As previous scholarship has detailed, base-rates in the difficulty of voting were probably reflective of a white majority feeling threatened by non-white political power. Put differently, the 2020 COVI probably picked up on salient differences among less-white states that were correlated with racial threat. The states with the higher COVIs in 2020, therefore, were more susceptible to feeling threatened by the results of the 2020 election, leading to racial backlash. In short: theories of racial threat and white backlash do a far better job of explaining the observed patterns in the 2021 legislative context than do arguments about shoring up overly-permissive voting regimes.

## District-Level Legislative Behavior

Having determined that the geographic concentration of the restrictive voting bills in 2021 was highly consistent with theories of racial threat, I turn to the sub-state level to better understand these processes. As discussed above, I expect to find evidence of backlash against racial threat in the whitest parts of the least white states.

Table \ref{tab:chamb} presents the results of OLS regressions investigating the likelihood that a legislator from a given district sponsored at least one restrictive (Models 1--4) or expansive (Models 5--8) voting provision in 2021.^[The ACS is missing demographic data on a handful of small legislative districts, explaining the variability in the observation count. However, the share of the population living in upper and lower chambers without demographic data is just 0.5% and 1.1%, respectively.] About 21% of lower chamber legislators and 20% of upper chamber legislators sponsored restrictive provisions; these numbers are 33% and 31% for expansive provisions, respectively.

\newpage
\begin{landscape}
\input{"../temp/district_reg.tex"}
\end{landscape}

As before, I plot the predicted probability of bill sponsorship for provisions to aid in the interpretation of the substantive relationships. Figure \ref{fig:mef-dis} presents like predicted likelihood of the sponsorship of a provision at different levels of district- and state-level racial compositions. Once again, all other covariates are held at their means.

```{r mefs-cha, echo = FALSE, message = F, warning = F, fig.width = 6, fig.height = 7, fig.cap = "\\label{fig:mef-dis}Provisions Sponsored by Chamber", fig.pos = "!ht", fig.align = "center"}
print(readRDS("temp/mef_chamb_good.rds"))
```

Table \ref{tab:chamb} and Figure \ref{fig:mef-dis} make a number of things immediately clear. Firstly, it is apparent that in both upper and lower legislative chambers, legislators from whiter districts are *more* likely to introduce voting bills with restrictive provisions and *less* likely to introduce expansive voting bills, ceteris paribus. Although Table \ref{tab:chamb} indicates that this relationship between district-level demographics is moderated by the share of the state that is white, Figure \ref{fig:mef-dis}(B) demonstrates that the substantive relationship is really very similar: the slope of the line in less-white states is effectively the same as that in the whiter states.

Figure \ref{fig:mef-dis}(A), however, provides evidence that strongly supports H3. While legislators from less-white districts in less-white states are among the *least* likely to sponsor restrictive bills, the legislators from the whitest parts of these same states are among the *most* likely to do so.

# Discussion

While legislators claim to pass restrictive voter policies under the guise of improved security, the data unambiguously points to white backlash to perceived racial threat. While discussing what she calls the "political work of fraud allegations," Lorraine Minnite argued a decade ago that claims of fraud could be explained by "the existence of marginalized subjects within the political culture whose presence alone stands in as the evidence of the alleged fraud" [@Minnite2010, 87]. In other words, the rhetoric of voter fraud draws boundaries around who does and does not count as a citizen, and whose political participation is inherently suspect.

A similar pattern emerges in this study of regressive voting laws introduced and passed in 2021, following the hard-fought and deeply polarizing 2020 presidential election and subsequent violent attempt to overturn the results on January 6th, 2021. Although the push to restrict voting access was extremely broad in 2021---more than 1 in 3 Americans lived in a district represented by a legislator who sponsored at least one of the 812 restrictive provisions---the push was concentrated in states and legislative districts with particular, systematic demographic characteristics.

\newpage
# References {-}
